% Govhack 2017 data exploration
% Ashley Noel Hinton
% `r format(Sys.time(), "%d %B %Y")`

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r libraries}
library(knitr)
library(rmarkdown)
library(stringi)
```

The day before Govhack started I was standing in a shop looking out
onto a busy Karangahape Road intersection, talking to the shop owner
about how there were lots of traffic accidents on the
intersection. Mostly they were trying to convince an employee to cross
at the pedestrian crossing rather than shooting between parked
cars. Someone suggested that surely information about how many
accidents happened at that intersection must be online somewhere, and
they searched on Google to find it. They couldn't find it.

This hack attempts to at least answer the question: how many traffic
accidents happen at this intersection?

## Data source

I will use the
[disaggregated crash data](https://www.nzta.govt.nz/safety/safety-resources/road-safety-information-and-tools/disaggregated-crash-data/) 
from NZ Transport Agency (NZTA) to try to answer this question. The data set has
been downloaded and is in this repo as a 62MB CSV file,
[Disaggregated-crash-data.csv](Disaggregated-crash-data.csv).
An accompanying metadata file has also been downloaded from NZTA,
[disaggregated-crash-data-metadata.csv](disaggregated-crash-data-metadata.csv).

```{r loadData}
crashData = read.csv(file = "Disaggregated-crash-data.csv")
summary(crashData)
dim(crashData)
```
Excellent, `r dim(crashData)[1]` rows in `r dim(crashData)[2]` columns
of data. What do the variables represent?

```{r loadMetaData}

## this file is *NOT* in UTF-8. best guess at encoding:
(encoding = stringi::stri_enc_detect(
    paste(readLines("disaggregated-crash-data-metadata.csv"),
          collapse = "\n"))[[1]]$Encoding[1])
## read the data using this encoding
crashMetadata = read.csv(file = "disaggregated-crash-data-metadata.csv",
                         stringsAsFactors = FALSE,
                         fileEncoding = toupper(encoding))
names(crashMetadata)

## what's in columns 3:5?
head(crashMetadata[, 3:5])
## nothing?
all(is.na((crashMetadata)[, 3:5]))
## drop them
crashMetadata = crashMetadata[, 1:2]

```

```{r printMetadata}

## print the table
kable(crashMetadata)

```

